---
title: "Practicum 1 - Bird Strikes"
author:
- Rakshika Raju, raju.r@northeastern.edu
- Pranith Rao Nayeneni, nayeneni.p@northeastern.edu
output:
  html_document:
    df_print: paged
---

### Loading MySQL package and connecting to MySQL database
```{r echo=FALSE}
# 1. Library
install.packages("RMySQL")
require(RMySQL)

# 2. Settings
db_user <- 'admin'
db_password <- 'password'
db_name <- 'bird_strikes'
db_host <- 'localhost'
db_port <- 3306

# 3. Read data from db
mydb <-  dbConnect(MySQL(), user = db_user, password = db_password,
                 dbname = db_name, host = db_host, port = db_port)
```

### Dropping all tables if they already exist
```{sql connection=mydb}
DROP TABLE IF EXISTS incident;
```

```{sql connection=mydb}
DROP TABLE IF EXISTS flight_phase;
```

```{sql connection=mydb}
DROP TABLE IF EXISTS aircraft;
```

```{sql connection=mydb}
DROP TABLE IF EXISTS airline;
```

```{sql connection=mydb}
DROP TABLE IF EXISTS weather_condition;
```

```{sql connection=mydb}
DROP TABLE IF EXISTS airport;
```

### 1
### Creating the database schema
```{sql connection=mydb}
CREATE TABLE airport (
aid int NOT NULL AUTO_INCREMENT,
airportName VARCHAR(255) NOT NULL DEFAULT "Unknown",
airportCode VARCHAR(255) NOT NULL DEFAULT "Unknown",
state VARCHAR(255) NOT NULL DEFAULT "Unknown",
PRIMARY KEY (aid)
);
```

#### Note: Using the table name as weather_condition as 'condition' is a keyword
```{sql connection=mydb}
CREATE TABLE weather_condition (
cid int NOT NULL AUTO_INCREMENT,
`condition` TEXT NOT NULL,
explanation VARCHAR(255) DEFAULT "None",
PRIMARY KEY (cid)
);
```

```{sql connection=mydb}
CREATE TABLE airline (
airlineID int NOT NULL AUTO_INCREMENT,
airlineName TEXT NOT NULL,
PRIMARY KEY (airlineID)
);
```

```{sql connection=mydb}
CREATE TABLE aircraft (
aircraftID int NOT NULL AUTO_INCREMENT,
makeModel TEXT NOT NULL,
PRIMARY KEY (aircraftID)
);
```

```{sql connection=mydb}
CREATE TABLE flight_phase (
flightPhaseID INT NOT NULL AUTO_INCREMENT,
flightPhaseName TEXT NOT NULL,
PRIMARY KEY (flightPhaseID)
)
```

```{sql connection=mydb}
CREATE TABLE incident (
iid INT NOT NULL,
`date` DATE,
origin INT NOT NULL,
airline INT NOT NULL,
aircraft INT NOT NULL,
flightPhase INT NOT NULL,
impact BOOLEAN NOT NULL DEFAULT FALSE,
cond INT NOT NULL,
PRIMARY KEY (iid),
FOREIGN KEY (origin) REFERENCES airport (aid),
FOREIGN KEY (airline) REFERENCES airline (airlineID),
FOREIGN KEY (aircraft) REFERENCES aircraft (aircraftID),
FOREIGN KEY (flightPhase) REFERENCES flight_phase (flightPhaseID),
FOREIGN KEY (cond) REFERENCES weather_condition (cid)
);
```

### 2
### Installing and loading the 'sqldf' package
```{r}
install.packages("sqldf")
library(sqldf)
```

```{r}
options(sqldf.driver = "SQLite")
```

### Loading the CSV file to a data frame
```{r}
df.raw <- read.csv("BirdStrikesData.csv", header = TRUE, stringsAsFactors = FALSE) 

# Special character is getting inserted by default from CSV so correcting the name
names(df.raw)[names(df.raw) == "ï..Record.ID"] <- "Record.ID"
```

### Creating the dataframe for airport table
```{r}
airport.df <- df.raw[, c("Airport..Name", "Origin.State")]
unknown.row <- c("", "")
airport.df <- rbind(airport.df, unknown.row)

df.airport <- sqldf("select 1 as aid, `Airport..Name` as airportName, `Origin.State` as state from `airport.df` group by `Origin.State`, `Airport..Name`")

n.airport <- nrow(df.airport)
df.airport[,1] <- seq(1, n.airport)
df.airport[1,2] <- "Unknown"
df.airport[1,3] <- "Unknown"
df.airport 
```

### Creating the dataframe for weather_condition table 
```{r}
df.condition <- sqldf("select 1 as cid, `Conditions..Sky` as condition from `df.raw` group by `Conditions..Sky`")
n.condition <- nrow(df.condition)
df.condition[,1] <- seq(1, n.condition)
df.condition[, "explanation"] = NA
df.condition
```

### Creating the dataframe for airline table 
```{r}
df.airline2 <- sqldf("select 1 as airlineID, `Aircraft..Airline.Operator` as airlineName from `df.raw` group by `Aircraft..Airline.Operator`")
df.airline <- df.airline2[!(df.airline2$airlineName=="MILITARY"),]
n.airline <- nrow(df.airline)
df.airline[,1] <- seq(1, n.airline)
df.airline[1,2] <- "Unknown"
df.airline
```


### Creating the dataframe for aircraft table 
```{r}
df.aircraft <- sqldf("select 1 as aircraftID, `Aircraft..Make.Model` as makeModel from `df.raw` group by `Aircraft..Make.Model`")
n.aircraft <- nrow(df.aircraft)
df.aircraft[,1] <- seq(1, n.aircraft)
df.aircraft[1,2] <- "Unknown"
df.aircraft
```

### Creating the dataframe for flight_phase table 
```{r}
flightPhaseID <- c(1, 2, 3, 4)
flightPhaseName <- c("unknown", "takeoff", "landing", "inflight")
df.fp <- data.frame(flightPhaseID, flightPhaseName)
df.fp
```


### Creating the dataframe for incidents table 
#### Removing the rows where date is empty or when the airline is a military airline
```{r}
df.incidents <- sqldf("select `Record.ID` as iid, `FlightDate` as `date`, `Airport..Name` as origin, `Origin.State` as originState, `Aircraft..Airline.Operator` as airline, `Aircraft..Make.Model` as aircraft, `When..Phase.of.flight` as flightPhase, `Effect..Indicated.Damage` as impact, `Conditions..Sky` as cond from `df.raw` where `date` IS NOT \"\" AND airline NOT IN (\"MILITARY\", \"Military\", \"military\")")

df.incidents$date <- as.Date(df.incidents$date, format = "%m/%d/%Y")
```

#### Modifying the values of the column 'impact' to be a boolean
```{r}
df.incidents$impact[df.incidents$impact == "Caused damage"] <- 1
df.incidents$impact[df.incidents$impact == "No damage"] <- 0
```

#### Harmonizing the flight phases to be one of landing(3), takeoff(2), inflight(4) and unknown(1).
```{r}
df.incidents$flightPhase[df.incidents$flightPhase == "Approach"] <- 3
df.incidents$flightPhase[df.incidents$flightPhase == "Descent"] <- 3
df.incidents$flightPhase[df.incidents$flightPhase == "Landing Roll"] <- 3
df.incidents$flightPhase[df.incidents$flightPhase == "Taxi"] <- 3
df.incidents$flightPhase[df.incidents$flightPhase == "Climb"] <- 2
df.incidents$flightPhase[df.incidents$flightPhase == "Parked"] <- 2
df.incidents$flightPhase[df.incidents$flightPhase == "Take-off run"] <- 2
df.incidents$flightPhase[df.incidents$flightPhase == ""] <- 1
```

```{r}
df.incidents.copy <- df.incidents
```

#### Creating foreign key relationships between the incident table and the other tables by replacing the values with their corresponding ids.
```{r}
n.incidents <- nrow(df.incidents.copy)
for (r in 1:n.incidents) {

a <- df.airport$aid[which(df.airport$`airportName` == df.incidents.copy$origin[r] & df.airport$`state` == df.incidents.copy$originState[r])]
if(length(a) == 0){
df.incidents$origin[r] <- 1;
} else {
df.incidents$origin[r] <- a;
}

b <- df.airline$airlineID[which(df.airline$`airlineName` == df.incidents.copy$airline[r])]
if(length(b) == 0){
df.incidents$airline[r] <- 1;
} else {
df.incidents$airline[r] <- b;
}

c <- df.aircraft$aircraftID[which(df.aircraft$`makeModel` == df.incidents.copy$aircraft[r])]
if(length(c) == 0){
df.incidents$aircraft[r] <- 1;
} else {
df.incidents$aircraft[r] <- c
}

d <- df.condition$cid[which(df.condition$`condition` == df.incidents.copy$cond[r])]
if(length(d) == 0){
df.incidents$cond[r] <- 1;
} else {
df.incidents$cond[r] <- d
}
}
```

#### Removing extra data
```{r}
df.incidents = subset(df.incidents, select = -c(originState) )
df.incidents
```


```{r}
dbSendQuery(mydb, "SET GLOBAL local_infile = true;")
```
### Inserting the 'airport' table into the database
```{r}
dbWriteTable(mydb, value = df.airport, name = "airport", append = TRUE, row.names = FALSE) 
```
### Inserting the 'aircraft' table into the database
```{r}
dbWriteTable(mydb, value = df.aircraft, name = "aircraft", append = TRUE, row.names = FALSE) 
```
### Inserting the 'weather_condition' table into the database
```{r}
dbWriteTable(mydb, value = df.condition, name = "weather_condition", append = TRUE, row.names = FALSE) 
```
### Inserting the 'airline' table into the database
```{r}
dbWriteTable(mydb, value = df.airline, name = "airline", append = TRUE, row.names = FALSE) 
```
### Inserting the 'flight_phase' table into the database
```{r}
dbWriteTable(mydb, value = df.fp, name = "flight_phase", append = TRUE, row.names = FALSE) 
```
### Inserting the 'incident' table into the database
```{r}
dbWriteTable(mydb, value = df.incidents, name = "incident", append = TRUE, row.names = FALSE) 
```

### 3
### Displaying the inserted tables.

```{sql connection=mydb}
SELECT * FROM airport LIMIT 20;
```

```{sql connection=mydb}
SELECT * FROM aircraft LIMIT 20;
```

```{sql connection=mydb}
SELECT * FROM weather_condition;
```

```{sql connection=mydb}
SELECT * FROM airline LIMIT 20;
```

```{sql connection=mydb}
SELECT * FROM flight_phase;
```

```{sql connection=mydb}
SELECT * FROM incident LIMIT 20;
```

### 4
### SQL query to find the number of bird strike incidents for each flight phase 
```{sql connection=mydb}
SELECT flight_phase.flightPhaseName, count(*) AS Average from incident, flight_phase 
WHERE incident.flightPhase = flight_phase.flightPhaseID 
GROUP BY incident.flightPhase;
```



### Average for reference
```{sql connection=mydb}
SELECT count(*) / (SELECT count(*) FROM flight_phase) AS TotalAverage FROM incident;
```

### 5
### SQL Query to find the flight phases that have above average bird strike incident.
```{sql connection=mydb}
 SELECT A.flightPhase, flight_phase.flightPhaseName, A.Average FROM (
 SELECT flightPhase, count(*) AS Average, (count(*) > ( SELECT count(*) / (SELECT count(*) FROM flight_phase) FROM incident)) AS greater
 FROM incident 
 GROUP BY incident.flightPhase) A, flight_phase
 WHERE A.greater = 1 AND A.flightPhase = flight_phase.flightPhaseID;
```

### 6
### SQL Query to find the average number of bird strike incidents by month (across all years)
```{sql connection=mydb}
SELECT month(date) AS Month, count(*)/12 AS Average 
FROM incident 
GROUP BY month(date) 
ORDER BY month(date);
```


### 7
### Column chart to visualize the number of bird strike incidents per year from 2005 to 2011.
```{sql connection=mydb, output.var="numberOfIncidents"}
SELECT YEAR(`date`) as incidentYear
FROM bird_strikes.incident 
WHERE YEAR(`date`) BETWEEN 2005 AND 2011
```

```{r}
barplot(table(numberOfIncidents),
main="Number of Bird Strike Incidents per year between 2005 and 2011",
xlab="Year",
ylab="Number of Incidents",
border="red",
col="blue",
density=10
)
```

### 8
### Stored procedure to add new bird strike incident to the database. 
```{sql connection=mydb}
DROP PROCEDURE IF EXISTS insert_incident;
```

```{sql connection=mydb}
Create PROCEDURE insert_incident(IN id int, IN dateOfIncident DATE, IN airportNameNew TEXT, IN airportStateNew TEXT, IN airlineNew TEXT, IN aircraftNew TEXT, IN flightPhaseNew TEXT, IN impactNew BOOLEAN, IN condNew TEXT)
BEGIN

SET @airlineValue = (SELECT airlineID FROM airline WHERE airlineName = airlineNew);
SET @aircraftValue = (SELECT aircraftID FROM aircraft WHERE makeModel = aircraftNew);
SET @flightPhaseValue = (SELECT flightPhaseID FROM flight_phase WHERE flightPhaseName = flightPhaseNew);
SET @condValue = (SELECT cid FROM weather_condition WHERE `condition` = condNew);
SET @airportID = (SELECT aid FROM airport WHERE airportName = airportNameNew AND state = airportStateNew);

IF @airportID IS NOT NULL THEN
INSERT INTO incident (iid, `date`, origin, airline, aircraft, flightPhase, impact, cond) VALUES (id, dateOfIncident, @airportID, @airlineValue, @aircraftValue, @flightPhaseValue, impactNew, @condValue);
ELSE
INSERT INTO airport (airportName, airportCode, state) VALUES (airportNameNew, "Unknown", airportStateNew);
SET @airportIDNew = (SELECT aid FROM airport WHERE airportName = airportNameNew AND state = airportStateNew);
INSERT INTO incident (iid, `date`, origin, airline, aircraft, flightPhase, impact, cond) VALUES (id, dateOfIncident, @airportIDNew, @airlineValue, @aircraftValue, @flightPhaseValue, impactNew, @condValue);
END IF;

END
```

### Testing 'insert_incident' stored procedure for an existing airport
```{sql connection=mydb}
CALL insert_incident(18101998, "2008/10/18", "BESSEMER ARPT", "Alabama", "AMERICA WEST AIRLINES", "B-737-200", "landing", FALSE, "Overcast");
```

```{sql connection=mydb}
SELECT * FROM incident where iid = 18101998;
```


### Testing 'insert_incident' stored procedure for a non existing airport
```{sql connection=mydb}
CALL insert_incident(19101998, "2010/10/19", "NEU AIRPORT", "MA", "AMERICA WEST AIRLINES", "B-737-200", "landing", FALSE, "Overcast");
```

#### Checking if the new airport is added to the 'airport' table
```{sql connection=mydb}
SELECT * FROM airport where airportName = "NEU AIRPORT";
```

```{sql connection=mydb}
SELECT * FROM incident where iid = 19101998;
```

### Disconnecting from the database
```{r}
dbDisconnect(mydb)
```


