---
title: "Practicum 2- Part 2 and Part 3"
author:
- Rakshika Raju, raju.r@northeastern.edu
- Pranith Rao Nayeneni, nayeneni.p@northeastern.edu
output: html_notebook
---

## Part 2

### Loading MySQL package and connecting to MySQL database
```{r echo=FALSE}
# 1. Library
#install.packages("RMySQL")
#install.packages("sqldf")
require(RMySQL)
require(sqldf)
options(sqldf.driver = "SQLite")

# 2. Settings
db_user <- 'admin'
db_password <- 'password'
db_name <- 'olap_pubmed'
db_host <- 'localhost'
db_port <- 3306

# 3. Read data from db
mydb <-  dbConnect(MySQL(), user = db_user, password = db_password,
                 dbname = db_name, host = db_host, port = db_port)
```

### Installing and loading of RSQLite package
```{r}
#install.packages("RSQLite")
library(RSQLite)
```

### Connecting to the SQLite database
```{r}
dbcon <- dbConnect(RSQLite::SQLite(), "PubMed.db")
```

```{sql connection=dbcon}
PRAGMA foreign_keys = ON;
```

### Dropping all tables if they already exist
```{sql connection=mydb}
DROP TABLE IF EXISTS fact_author;
```

### Creating the table
```{sql connection=mydb}
CREATE TABLE fact_author (
aid INT NOT NULL,
first_name TEXT NOT NULL,
last_name TEXT NOT NULL,
num_of_articles_published INT NOT NULL,
avg_articles_per_year DECIMAL(5, 2) NOT NULL
);
```


### Getting data for inserting into the fact table
```{sql connection=dbcon, output.var="authordata"}
SELECT Author.aid AS aid, Author.first_name AS first_name , Author.last_name AS last_name, COUNT(Author_Article.aid) AS num_of_articles_per_year, Article.article_year as Year
    FROM Author
         INNER JOIN Author_Article ON Author_Article.aid = Author.aid
         INNER JOIN Article ON Author_Article.article_id = Article.article_id
GROUP BY Author_Article.aid, Article.article_year;
```

### Manipulating data in R to get the correct format
```{r}
authdf <- sqldf("SELECT aid, first_name, last_name, SUM(num_of_articles_per_year) as num_of_articles_published,  ROUND(CAST(SUM(num_of_articles_per_year) AS REAL)/(MAX(Year) - MIN(YEAR) + 1), 2) as avg_articles_per_year FROM authordata GROUP BY aid")
```
```{r}
print(authdf)
```

### Insert rows from the data frame to the table
```{r}
for (i in 1:nrow(authdf)){
sqlcmd <- paste0("INSERT INTO fact_author(aid, first_name, last_name, num_of_articles_published, avg_articles_per_year) VALUES (",authdf[i, 1],",","\"",authdf[i, 2],"\"",",","\"",authdf[i, 3],"\"",",",authdf[i, 4],",",authdf[i, 5],")")
rs <- dbExecute(mydb, sqlcmd)
}
```

### Displaying the fact table
```{sql connection=mydb}
SELECT aid, first_name, last_name, num_of_articles_published, avg_articles_per_year FROM fact_author;
```

## Part 3

### Top 10 authors in terms of number of articles published
```{sql connection=mydb}
SELECT aid, first_name, last_name, num_of_articles_published FROM fact_author 
ORDER BY num_of_articles_published DESC
LIMIT 10;
```

### Disconnecting from database

```{r}
dbDisconnect(mydb)
```